`You Draw It': Implementation of visually fitted trends with \texttt{r2d3}

### 2022-07-01 2022-sdss-you-draw-it-manuscript.Rmd
rough rough draft

When completing the graphical task, users are shown an interactive scatter-plot (Figure 1) along with the prompt, "Use your mouse to fill in the trend in the yellow box region." The yellow box region moves along as the user draws their trend-line, providing a visual cue which indicates where the user still needs to complete a trend line. After the entire domain has been visually estimated or predicted, the yellow region disappears, indicating the participant has completed the task. Prior to study participation, example gifs are shown and users are asked to complete practice plots to ease the learning curve of the task. Visit \url{emily-robinson.shinyapps.io/can-you-draw-it} for a test applet.

@@

Parameters for aesthetic design choices are defined in a list of options and `r2d3` passes these to the D3.js code. For instance, we can specify the buffer space allowed for the $x$ and $y$ axes to avoid users to anchor their lines to the axes limits. For `D3.js` source code, visit \url{https://github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/main-d3v5.js}.

@@

We conducted a study in order to validate 'You Draw It' as a method for graphical testing, comparing results to the less technological method utilized in Mosteller et al. \shortcite{mosteller1981eye}. Results from our study were consistent with those found in the previous study; when shown points following a linear trend, participants tended to fit the slope of the first principal component over the slope of the ordinary least-squares regression line (Figure 3). This trend was most prominent when shown data simulated with larger variances. This study reinforces the differences between intuitive visual model fitting and statistical model fitting, providing information about human perception as it relates to the use of statistical graphics.

@@

Feedback data from conducted studies are collected and stored in a database for analysis (Figure 4). Within the collected feedback data, we know the simulated data points, the predicted values from the statistical model, and the predicted values from the user drawn line. In our initial studies, a unique data set was simulated independently for each participant. Therefore, we evaluate the accuracy of the user drawn line by observing the deviation, vertical residuals, between the user drawn line and the predicted values from the statistical model. We use a Generalized Additive Mixed Model (GAMM) to model the vertical residuals in order to statistically compare visually fitted trends to actual metrics, simulated data models, or statistical regression results. A benefit of using a GAMM is the estimation of smoothing splines, allowing for flexibility in the residual trend.

\begin{figure}[ht]
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{images/eyefitting-trial-plot}}
\caption{Example of three trend lines showing the the OLS fitted, PCA fitted, and participant drawn values overlaid on the simulated data points}
\label{eyefitting-trial-plot}
\end{center}
\end{figure}

\begin{figure}[ht]
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{images/eyefitting-lmer-plot}}
\caption{Estimated trends of residuals (vertical deviation of participant drawn points from both the OLS (blue) and PCA (orange) fitted points) as fit by a linear mixed model.}
\label{eyefitting-lmer-plot}
\end{center}
\end{figure}

\begin{figure}[ht]
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{images/eyefitting-gamm-plot}}
\caption{Estimated trends of residuals (vertical deviation of participant drawn points from both the OLS (blue) and PCA (orange) fitted points) as fit by a generalized additive mixed model.}
\label{eyefitting-gamm-plot}
\end{center}
\end{figure}

@@

In this work, we implemented and validated ‘You Draw It’ as a way to measure the patterns we see in data. We demonstrated the use of generalized additive models to statistically model participant data. While technical details of the development process are presented here, we intend to create an R package designed for easy implementation of 'You Draw It' task plots in order to make this tool accessible to other researchers. Further investigation is necessary to implement this method real data in order to facilitate scientific communication.

>>>>

When completing the graphical task, users are shown an interactive scatter-plot (Figure 1) along with the prompt, "Use your mouse to fill in the trend in the yellow box region." The yellow box region moves along as the user draws their trend-line, providing a visual cue which indicates where the user still needs to complete a trend line. After the entire domain has been visually estimated or predicted, the yellow region disappears, indicating the participant has completed the task. Prior to study participation, example gifs are shown and users are asked to complete practice plots to ease the learning curve of the task. ^[Visit \url{emily-robinson.shinyapps.io/can-you-draw-it} for a test applet.]

@@

Parameters for aesthetic design choices are defined in a list of options and `r2d3` passes these to the D3.js code. For instance, we can specify the buffer space allowed for the $x$ and $y$ axes to avoid users to anchor their lines to the axes limits. ^[For `D3.js` source code, visit \url{https://github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/main-d3v5.js}.]

@@

We conducted a study in order to validate 'You Draw It' as a method for graphical testing, comparing results to the less technological method utilized in Mosteller et al. \shortcite{mosteller1981eye}. Results from our study were consistent with those found in the previous study; when shown points following a linear trend, participants tended to fit the slope of the first principal component over the slope of the ordinary least-squares regression line (Figure 3). This trend was most prominent when shown data simulated with larger variances. This study reinforced the differences between intuitive visual model fitting and statistical model fitting, providing information about human perception as it relates to the use of statistical graphics.

@@

\begin{figure}[ht]
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{images/eyefitting-trial-plot}}
\caption{Example of validation feedback data where three trend lines show the the OLS fitted, PCA fitted, and participant drawn values overlaid on the simulated data points}
\label{eyefitting-trial-plot}
\end{center}
\end{figure}

Feedback data from conducted studies were collected and stored in a database for analysis (Figure 4). Within the collected feedback data, we knew the simulated data points, the predicted values from the statistical model, and the predicted values from the user drawn line. In our initial studies, a unique data set was simulated independently for each participant. Therefore, we evaluate the accuracy of the user drawn line by observing the deviation, vertical residuals, between the user drawn line and the predicted values from the statistical model. Due to participant variation, we use mixed models to evaluate the vertical residuals in order to statistically compare visually fitted trends to actual metrics, simulated data models, or statistical regression results. In our validation study, we first analyzed the residual trend using a linear mixed model (LMM), thus constraining the fit to a linear trend (Figure 5). In addition, we fit a Generalized Additive Mixed Model (GAMM) to allow for flexibility in the residual trend and to generalized the method to nonlinear models (Figure 6).

\begin{figure}[ht]
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{images/eyefitting-lmer-plot}}
\caption{Validation study estimated trends of residuals (vertical deviation of participant drawn points from both the OLS (blue) and PCA (orange) fitted points) as fit by a linear mixed model.}
\label{eyefitting-lmer-plot}
\end{center}
\end{figure}

\begin{figure}[ht]
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{images/eyefitting-gamm-plot}}
\caption{Validation study estimated trends of residuals (vertical deviation of participant drawn points from both the OLS (blue) and PCA (orange) fitted points) as fit by a generalized additive mixed model.}
\label{eyefitting-gamm-plot}
\end{center}
\end{figure}

After validating the 'You Draw It' method, we conducted a study utilizing the new method to evaluate participants ability to make forecasts for exponentially increasing data on log and linear scales. Along with analyzing the feedback data with the GAMM method for flexibility due to nonlinear data, we used spaghetti plots to visualize participants forecasts compared to the nonlinear least squares statistical model and make comparisons between two chart design features (Figure 6).

@@

In this work, we implemented and validated ‘You Draw It’ as a new method to measure the patterns we see in data. We demonstrated the use of generalized additive models to statistically model participant feedback data and generalized the analysis method to nonlinear trends. While technical details of the development process are presented here, we intend to create an R package designed for easy implementation of 'You Draw It' task plots in order to make this tool accessible to other researchers. Further investigation is necessary to implement this method with real data in order to facilitate scientific communication.

### 2022-07-01 2022-sdss-you-draw-it-manuscript.tex
rough rough draft

When completing the graphical task, users are shown an interactive
scatter-plot (Figure 1) along with the prompt, ``Use your mouse to fill
in the trend in the yellow box region.'' The yellow box region moves
along as the user draws their trend-line, providing a visual cue which
indicates where the user still needs to complete a trend line. After the
entire domain has been visually estimated or predicted, the yellow
region disappears, indicating the participant has completed the task.
Prior to study participation, example gifs are shown and users are asked
to complete practice plots to ease the learning curve of the task. Visit
\url{emily-robinson.shinyapps.io/can-you-draw-it} for a test applet.

@@

Parameters for aesthetic design choices are defined in a list of options
and \texttt{r2d3} passes these to the D3.js code. For instance, we can
specify the buffer space allowed for the \(x\) and \(y\) axes to avoid
users to anchor their lines to the axes limits. For \texttt{D3.js}
source code, visit
\url{https://github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/main-d3v5.js}.

@@

We conducted a study in order to validate `You Draw It' as a method for
graphical testing, comparing results to the less technological method
utilized in Mosteller et al.~\shortcite{mosteller1981eye}. Results from
our study were consistent with those found in the previous study; when
shown points following a linear trend, participants tended to fit the
slope of the first principal component over the slope of the ordinary
least-squares regression line (Figure 3). This trend was most prominent
when shown data simulated with larger variances. This study reinforces
the differences between intuitive visual model fitting and statistical
model fitting, providing information about human perception as it
relates to the use of statistical graphics.

@@

Feedback data from conducted studies are collected and stored in a
database for analysis (Figure 4). Within the collected feedback data, we
know the simulated data points, the predicted values from the
statistical model, and the predicted values from the user drawn line. In
our initial studies, a unique data set was simulated independently for
each participant. Therefore, we evaluate the accuracy of the user drawn
line by observing the deviation, vertical residuals, between the user
drawn line and the predicted values from the statistical model. We use a
Generalized Additive Mixed Model (GAMM) to model the vertical residuals
in order to statistically compare visually fitted trends to actual
metrics, simulated data models, or statistical regression results. A
benefit of using a GAMM is the estimation of smoothing splines, allowing
for flexibility in the residual trend.

\begin{figure}[ht]
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{images/eyefitting-trial-plot}}
\caption{Example of three trend lines showing the the OLS fitted, PCA fitted, and participant drawn values overlaid on the simulated data points}
\label{eyefitting-trial-plot}
\end{center}
\end{figure}

\begin{figure}[ht]
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{images/eyefitting-lmer-plot}}
\caption{Estimated trends of residuals (vertical deviation of participant drawn points from both the OLS (blue) and PCA (orange) fitted points) as fit by a linear mixed model.}
\label{eyefitting-lmer-plot}
\end{center}
\end{figure}

\begin{figure}[ht]
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{images/eyefitting-gamm-plot}}
\caption{Estimated trends of residuals (vertical deviation of participant drawn points from both the OLS (blue) and PCA (orange) fitted points) as fit by a generalized additive mixed model.}
\label{eyefitting-gamm-plot}
\end{center}
\end{figure}

@@

In this work, we implemented and validated `You Draw It' as a way to
measure the patterns we see in data. We demonstrated the use of
generalized additive models to statistically model participant data.
While technical details of the development process are presented here,
we intend to create an R package designed for easy implementation of
`You Draw It' task plots in order to make this tool accessible to other
researchers. Further investigation is necessary to implement this method
real data in order to facilitate scientific communication.

>>>>

When completing the graphical task, users are shown an interactive
scatter-plot (Figure 1) along with the prompt, ``Use your mouse to fill
in the trend in the yellow box region.'' The yellow box region moves
along as the user draws their trend-line, providing a visual cue which
indicates where the user still needs to complete a trend line. After the
entire domain has been visually estimated or predicted, the yellow
region disappears, indicating the participant has completed the task.
Prior to study participation, example gifs are shown and users are asked
to complete practice plots to ease the learning curve of the
task.\footnote{Visit \url{emily-robinson.shinyapps.io/can-you-draw-it}
  for a test applet.}

@@

Parameters for aesthetic design choices are defined in a list of options
and \texttt{r2d3} passes these to the D3.js code. For instance, we can
specify the buffer space allowed for the \(x\) and \(y\) axes to avoid
users to anchor their lines to the axes limits.\footnote{For
  \texttt{D3.js} source code, visit
  \url{https://github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/main-d3v5.js}.}

@@

We conducted a study in order to validate `You Draw It' as a method for
graphical testing, comparing results to the less technological method
utilized in Mosteller et al.~\shortcite{mosteller1981eye}. Results from
our study were consistent with those found in the previous study; when
shown points following a linear trend, participants tended to fit the
slope of the first principal component over the slope of the ordinary
least-squares regression line (Figure 3). This trend was most prominent
when shown data simulated with larger variances. This study reinforced
the differences between intuitive visual model fitting and statistical
model fitting, providing information about human perception as it
relates to the use of statistical graphics.

@@

\begin{figure}[ht]
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{images/eyefitting-trial-plot}}
\caption{Example of validation feedback data where three trend lines show the the OLS fitted, PCA fitted, and participant drawn values overlaid on the simulated data points}
\label{eyefitting-trial-plot}
\end{center}
\end{figure}

Feedback data from conducted studies were collected and stored in a
database for analysis (Figure 4). Within the collected feedback data, we
knew the simulated data points, the predicted values from the
statistical model, and the predicted values from the user drawn line. In
our initial studies, a unique data set was simulated independently for
each participant. Therefore, we evaluate the accuracy of the user drawn
line by observing the deviation, vertical residuals, between the user
drawn line and the predicted values from the statistical model. Due to
participant variation, we use mixed models to evaluate the vertical
residuals in order to statistically compare visually fitted trends to
actual metrics, simulated data models, or statistical regression
results. In our validation study, we first analyzed the residual trend
using a linear mixed model (LMM), thus constraining the fit to a linear
trend (Figure 5). In addition, we fit a Generalized Additive Mixed Model
(GAMM) to allow for flexibility in the residual trend and to generalized
the method to nonlinear models (Figure 6).

\begin{figure}[ht]
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{images/eyefitting-lmer-plot}}
\caption{Validation study estimated trends of residuals (vertical deviation of participant drawn points from both the OLS (blue) and PCA (orange) fitted points) as fit by a linear mixed model.}
\label{eyefitting-lmer-plot}
\end{center}
\end{figure}

\begin{figure}[ht]
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{images/eyefitting-gamm-plot}}
\caption{Validation study estimated trends of residuals (vertical deviation of participant drawn points from both the OLS (blue) and PCA (orange) fitted points) as fit by a generalized additive mixed model.}
\label{eyefitting-gamm-plot}
\end{center}
\end{figure}

After validating the `You Draw It' method, we conducted a study
utilizing the new method to evaluate participants ability to make
forecasts for exponentially increasing data on log and linear scales.
Along with analyzing the feedback data with the GAMM method for
flexibility due to nonlinear data, we used spaghetti plots to visualize
participants forecasts compared to the nonlinear least squares
statistical model and make comparisons between two chart design features
(Figure 6).

@@

In this work, we implemented and validated `You Draw It' as a new method
to measure the patterns we see in data. We demonstrated the use of
generalized additive models to statistically model participant feedback
data and generalized the analysis method to nonlinear trends. While
technical details of the development process are presented here, we
intend to create an R package designed for easy implementation of `You
Draw It' task plots in order to make this tool accessible to other
researchers. Further investigation is necessary to implement this method
with real data in order to facilitate scientific communication.
